---
toc: true
title: "Simple Imputation -  malignant breast cancer by ER using SEER 12, 1992–2020"
author: "AMF"
format: html
editor: visual
date: "January 31, 2024"
date-modified: last-modified
---



#### Data Source
National Cancer Institute. Surveillance, Epidemiology, and End Results (SEER12) Program Populations (1992-2020). Available at: [www.seer.cancer.gov/popdata](https://www.seer.cancer.gov/popdata). Released November 2023. National Cancer Institute, DCCPS, Surveillance Research Program.

#### Morphology
#### Infiltrating Ductal Carcinoma (IDC)
:::{.callout-note}
-   **8500/3**: Infiltrating duct carcinoma, NOS'
-   **8523/3**: Infiltrating duct mixed with other types of carcinoma'
:::

ER Status Recode for Breast Cancer (1990+): 'Positive', 'Negative', 'Borderline/Unknown', 'Recode not available


#### load required libraries
```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary library for data fetching
library(readr)
library(ggplot2)
library(dplyr)
library(gridExtra)

```

##### Importing Encrypted SEER12 Data Files from GitHub Repository

Discuss the process involved in accessing locked data files from SEER12 that have been stored on GitHub.

```{r, message=FALSE, warning=FALSE}
# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/Incidence/see12/seer12_er.csv"
df <- read_csv(url)

```

```{r, message=FALSE, warning=FALSE, echo=FALSE}


#| label: fig1
#| fig-cap: "Figure 1. Trends in the prevalence of estrogen receptor (ER) status for invasive female breast cancer cases in the United States are depicted in two parts. Figure A presents the annual proportions of breast cancer cases with unspecified ER status. Figure B details the yearly proportions of breast cancers with confirmed ER status that are ER-negative."
#| 
names(df)<-c("age", "year", "er", "rate", "cases", "py")
#unique(df$er)
df$er[df$er==3]<-2 

# Aggregate cases by age, year, and er
df_a <- df %>%
  group_by(age, year, er) %>%
  summarise(total_cases = sum(cases), .groups = 'drop')

# Recode the race and stage values
df_a$er <- factor(df_a$er, levels = c(0, 1, 2),
                labels = c("Positive", "Negative", "Unknown"))

# Filter out rows with "Year of diagnosis" in the 'year' column and convert to numeric
#df <- subset(df, df$year != "Year of diagnosis")
df_a$year <- as.numeric(df_a$year) + 1991
#unique(df_a$year)
df1<-df_a
df2<-subset(df1, df1$age >=40)


# Sum the number of ER cases by year and age
df_summary <- df2 %>%
  group_by(year, age) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df3<-merge(df_summary, df2, by=c("age", "year"))
# Sum the number of ER cases by year and age
df_s <- df2 %>%
  group_by(year, er) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
# Sum the number of ER cases by year and age
df_s1 <- df2 %>%
  group_by(year) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df4<-merge(df_s, df_s1, by=c("year"))
# Proportion of unknown
df4$prop<-(df4$total_cases.x / df4$total_cases.y)
# Filter for 'Unknown' ER only
df_unknown_er <- df4 %>%
  filter(er == "Unknown")

df_unknown_er<-subset(df_unknown_er, df_unknown_er$year>=1992)

# Create a line plot in Lance style

a <- ggplot(df_unknown_er, aes(x = year, y = prop, group = 1)) +
  geom_line(color = "black", size = 1) +  # Change line color and size as needed
  ylim(0,0.5) +
  geom_point(color = "black", size = 2) + # Add points and customize their color and size as needed
  theme_minimal() +
  labs(
    title = "Year at diagnosis",
    x = "Year",
    y = "% of Unknown ER"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    text = element_text(size = 12),
    aspect.ratio = 2.5 # Example aspect ratio, adjust as needed
  )


## Proportion Age at diagnosis

df2a<-subset(df2, df2$year>=1992 & df2$year <=2008) #1991 represents 1992-2020

# Sum the number of ER cases by year and age
dfb <- df2a %>%
  group_by(age) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group

# Sum the number of ER cases by year and age
dfc <- df2a %>%
  group_by(age, er) %>%
  summarise(total_cases = sum(total_cases), .groups = 'drop') # Sum of 'cases' within each 'year' and 'age' group
df4a<-merge(dfc, dfb, by=c("age"))

# Proportion of unknown
df4a$prop<-(df4a$total_cases.x / df4a$total_cases.y)

# Filter for 'Unknown' ER only
df_age <- df4a %>%
  filter(er == "Negative")

# Create a line plot in Lance style
b <- ggplot(df_age, aes(x = age, y = prop, group = 1)) +
  geom_line(color = "black", size = 1) +  # Change line color and size as needed
  ylim(0,0.5) +
  geom_point(color = "black", size = 2) + # Add points and customize their color and size as needed
  theme_minimal() +
  labs(
    title = "Age at diagnosis",
    x = "Age",
    y = "% ER Negative"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 12),
    aspect.ratio = 2.5 # Set a specific aspect ratio to make the plot 'skinny'
  )


grid.arrange(a,b,
             ncol=2,
             top = grid::textGrob('', gp=grid::gpar(fontsize=5)))


```
Figure A presents the annual proportions of breast cancer cases with unknown ER status. Figure B details the yearly proportions of breast cancers with ER-negative status."


#### Proportion of unknown by ER status and breast cancer subtype.

```{r, message=FALSE, warning=FALSE, echo=FALSE}

library(readr)
library(ggplot2)
library(gridExtra)

# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/ITEB-breast-cancer/main/survival_article/data/seer12_proportion_unk_race_er.csv"
df <- read_csv(url)
names(df)<-c("year", "race", "er", "rate", "cases", "py")
df$year <- as.numeric(df$year) + 1991
# Check the result
df<-subset(df, df$year > 1991)
# Recode the race and stage values
df$race <- factor(df$race, levels = c(0, 1, 2, 3, 4, 5),
                    labels = c("NHW", "NHB", "AIAN", "API", "HIS", "Unknown"))
df$er[df$er==3]<-2
df$er <- factor(df$er, levels = c(0, 1, 2),
                  labels = c("Positive", "Negative", "Unknown"))

# Now, aggregate both cases and py by the new age groups as well as by race and subtype
data <- aggregate(cbind(cases, py) ~ race + year + er, data=df, sum)
# estimate the total number of er cases by year and race
estimated_cases <- aggregate(data$cases, by = list(data$race, data$year), sum)
names(estimated_cases)<-c("race", "year", "total")
df1<-merge(data, estimated_cases, by=c("race", "year"))
df1$prop<-(df1$cases / df1$total)
#Create a line plot in Lance style
df2<-subset(df1, df1$er == "Unknown")
df3<-subset(df2, df2$race== "NHW"|df2$race=="NHB"|df2$race=="API"|df2$race=="HIS")

a1<-ggplot(df3, aes(x = year, y = prop, group = race, color = race)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  ylim(0, 0.5) +
  scale_color_manual(values = c("NHW" = "darkblue", "NHB" = "darkred", "API" = "purple", "HIS" = "darkorange")) +
  labs(
    title = "ER",
    x = "Year",
    y = "% of Unknown ER"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 12),
    aspect.ratio = 2.5,
    legend.position = "bottom",
    legend.text = element_text(size = 5), # This will reduce the legend text size
    legend.title = element_text(size = 6), # This will reduce the legend title size
    legend.key.size = unit(0.5, "cm") # This will reduce the size of the colored keys
  )


# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/ITEB-breast-cancer/main/survival_article/data/seer12_proportion_ukn_race_subtype.csv"
df <- read_csv(url)

names(df)<-c("year", "race", "subtype", "rate", "cases", "py")
df$year <- as.numeric(df$year) + 1991
# Check the result
df<-subset(df, df$year > 1991)
# Recode the race and stage values
df$race <- factor(df$race, levels = c(0, 1, 2, 3, 4, 5),
                  labels = c("NHW", "NHB", "AIAN", "API", "HIS", "Unknown"))
#
df$subtype[df$subtype==5]<-4
df$subtype <- factor(df$subtype, levels = c(0, 1, 2, 3, 4),
                labels = c("HR+/HER2+", "HR-/HER2+", "HR+/HER2-", "HR-/HER2-", "Unknown"))

# Now, aggregate both cases and py by the new age groups as well as by race and subtype
data <- aggregate(cbind(cases, py) ~ race + year + subtype, data=df, sum)
# estimate the total number of er cases by year and race
estimated_cases <- aggregate(data$cases, by = list(data$race, data$year), sum)
names(estimated_cases)<-c("race", "year", "total")
df1<-merge(data, estimated_cases, by=c("race", "year"))
df1$prop<-(df1$cases / df1$total)

#Create a line plot in Lance style

df2<-subset(df1, df1$year >=2010 & df1$subtype == "Unknown")
df3<-subset(df2, df2$race== "NHW"|df2$race=="NHB"|df2$race=="API"|df2$race=="HIS")

b1<- ggplot(df3, aes(x = year, y = prop, group = race, color = race)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  ylim(0, 0.5) +
  scale_color_manual(values = c("NHW" = "darkblue", "NHB" = "darkred", "API" = "purple", "HIS" = "darkorange")) +
  labs(
    title = "Subtype",
    x = "Year",
    y = "% of Unknown Subtype"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    text = element_text(size = 12),
    aspect.ratio = 2.5,
    legend.position = "bottom",
    legend.text = element_text(size = 5), # This will reduce the legend text size
    legend.title = element_text(size = 6), # This will reduce the legend title size
    legend.key.size = unit(0.5, "cm") # This will reduce the size of the colored keys
  )

###
grid.arrange(a1,b1,
             ncol=2,
             top = grid::textGrob('', gp=grid::gpar(fontsize=5)))


```



##### Simple imputation for malignant breast cancer by ER

A simple imputation method to correct the rates of ER-positive and ER-negative was based on the method proposed by[Anderson at al., 2011](https://pubmed.ncbi.nlm.nih.gov/21753181/)

in brief:



:::{.callout-tip}


We developed a simple imputation model to correct the apparent rates of ER-positive and ER-negative breast cancer for missing ER data. For each age `a` and year `t`, we partitioned the observed total number of incident breast cancers $0^I_{at}$ according to ER status. Hence

$$
I^I_{at} = P^0_{at} + N^0_{at} + U^0_{at} \tag{1}
$$

in which $P^0_{at}$, $N^0_{at}$, and $U^0_{at}$ are the observed ER-positive, ER-negative, and unknown counts, respectively. Prior studies have analyzed $P^0_{at}$ and $N^0_{at}$, but if $U^0_{at}$ varies by either `a` or `t` these results may be biased. Our imputation method estimates the unobserved complete data

$$
I^0_{at} = P^C_{at} + N^C_{at} \tag{2}
$$

in which $P^C_{at}$ and $N^C_{at}$ are the true numbers of ER-positive and ER-negative counts, respectively.

Our model assumes that unknown ER status is missing at random within a single year of age `a` and calendar year `t` of diagnosis. Under this model, the observed probability $\pi_{at} = \dfrac{P^0_{at}}{P^0_{at} + N^0_{at}}$ on the basis of patients for whom we have complete information is an unbiased estimator of the true probability at the population level that an incident breast cancer diagnosed among women age `a` and calendar year `t` is ER positive. The equations

$$
P^C_{at} = \hat{\pi}_{at} \times I^0_{at} \tag{3}
$$
$$
N^C_{at} = (1 - \hat{\pi}_{at}) \times I^0_{at} \tag{4}
$$

provide unbiased estimators of the true numbers of ER-positive and ER-negative breast cancers in the population.
:::



```{r, message=FALSE, warning=FALSE, echo=FALSE}

# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/Incidence/see12/seer12_er.csv"
df <- read_csv(url)

names(df)<-c("age", "year", "er", "rate", "cases", "py")
df$er[df$er==3]<-2 

# Aggregate cases by age, year, and er
df_a <- df %>%
  group_by(age, year, er, py) %>%
  summarise(total_cases = sum(cases), .groups = 'drop')

# Recode the race and stage values
df_a$er <- factor(df_a$er, levels = c(0, 1, 2),
                  labels = c("Positive", "Negative", "Unknown"))

# Filter out rows with "Year of diagnosis" in the 'year' column and convert to numeric
#df <- subset(df, df$year != "Year of diagnosis")
df_a$year <- as.numeric(df_a$year) + 1991

df1<-df_a

df2<-subset(df1, df1$age >=40)


#
# Display the first few rows of the data to understand what we are working with


# Group data by age and year to calculate observed counts and probabilities
library(dplyr)

df2_summary <- df2 %>%
  group_by(age, year) %>%
  summarise(
    total_positive = sum(total_cases[er == "Positive"]),
    total_negative = sum(total_cases[er == "Negative"]),
    total_known = total_positive + total_negative,
    total_cases = sum(total_cases)
  ) %>%
  ungroup()

# Calculate the observed probabilities for known ER status
df2_summary <- df2_summary %>%
  mutate(
    Observed_Probability_Positive = total_positive / total_known,
    Observed_Probability_Negative = total_negative / total_known
  )

# Estimate the true counts for ER-positive and ER-negative cases using the observed probabilities
df2_summary <- df2_summary %>%
  mutate(
    Estimated_Positive = Observed_Probability_Positive * total_cases,
    Estimated_Negative = Observed_Probability_Negative * total_cases
  )

# Display the resulting estimates
df2_summary$Estimated_Positive<-round(df2_summary$Estimated_Positive, 0)
df2_summary$Estimated_Negative<-round(df2_summary$Estimated_Negative, 0)

# You may want to save the updated summary dataframe with estimations
#write.csv(df2_summary, 'df2_summary_with_estimations.csv', row.names = FALSE)



# Create a long format

df2_summary$Estimated_Positive<-as.numeric(df2_summary$Estimated_Positive)
df2_summary$Estimated_Negative<-as.numeric(df2_summary$Estimated_Negative)




#write.csv(df2_summary, "C:/Users/filhoam/Desktop/Breast/Incidence/imputation/imputed.csv")

# Read the dataset from the provided URL
url <- "https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/Incidence/imputation/imputed.csv"
dat <- read_csv(url)

# Assuming your data frame is named df2_summary
df2_aggregated_by_year <- dat %>%
  group_by(year) %>%
  summarise(
    total_positive = sum(total_positive),
    total_negative = sum(total_negative),
    Estimated_Positive = sum(Estimated_Positive, na.rm = TRUE), # Replace with actual column name
    Estimated_Negative = sum(Estimated_Negative, na.rm = TRUE)  # Replace with actual column name
  )


#head(df2_aggregated_by_year)


# Population
url <- "https://raw.githubusercontent.com/filhoalm/Breast_cancer/main/Incidence/imputation/py.csv"
py <- read_csv(url)
names(py)<-c("year", "rate", "counts", "py")
py$year <- as.numeric(py$year) + 1991
py1<-py[c(1,4)]

## Merge
df2_aggregated<-merge(df2_aggregated_by_year, py1, by=c("year"))

#

df2_aggregatedA<-df2_aggregated[c(1,2,6)]
df2_aggregatedA$index<-"total_positive"

df2_aggregatedB<-df2_aggregated[c(1,3,6)]
df2_aggregatedB$index<-"total_negative"
names(df2_aggregatedB)[2]<-c("total_positive")

df2_aggregatedC<-df2_aggregated[c(1,4,6)]
df2_aggregatedC$index<-"Estimated_Positive"
names(df2_aggregatedC)[2]<-c("total_positive")

df2_aggregatedD<-df2_aggregated[c(1,5,6)]
df2_aggregatedD$index<-"Estimated_Negative"
names(df2_aggregatedD)[2]<-c("total_positive")



df2_long<-rbind(df2_aggregatedA, df2_aggregatedB, df2_aggregatedC, df2_aggregatedD)
names(df2_long)[2]<-c("cases")


# rates

df2_long$rate<-(df2_long$cases / df2_long$py)*100000

## plot

long<-subset(df2_long, df2_long$year>=1992)

library(ggplot2)

# Create a line plot in a skinny Lance style
ggplot(long, aes(x = year, y = rate, group = index, color = index)) +
  geom_line(aes(linetype=index, color=index)) +
  geom_point(aes(shape = index)) + # Point size
  scale_color_manual(values = c("darkblue", "darkblue", "darkred", "darkred")) + # Replace with desired colors
  theme_minimal() + # Lance style typically implies a minimalistic theme
  labs(
    title = "",
    x = "Year of diagnosis",
    y = "Crude rates per 100,000"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    text = element_text(size = 12), # Change text size as needed
    aspect.ratio = 2.5 # Adjust this value to make the figure skinnier or wider
  )

head(dat)

```






